---
title: "Warden and Gore waterbird analyses"
author: "Adrian Pinder"
date and time: '`r Sys.time()`'
output: html_document
software: 'RStudio: Version 1.1.463 – © 2009-2018 RStudio, Inc. R version: `r getRversion()`'
editor_options: 
  chunk_output_type: console
---

Git repository https://github.com/AdrianMP62/Warden-2019  

RStudio: Version 1.1.463 – © 2009-2018 RStudio, Inc. R version: `r getRversion()`  
Date and time: `r Sys.time()`

Uses the following datafiles:  
*  
*  
*  

Date and time: `r Sys.time()`

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(include = FALSE)
```

```{r}
options(scipen=999)
```

```{r}
library(reshape2)
library(vegan)
library(ggplot2)
library(stringr)
library(knitr)
library(simba)
source("bio_env_ext.R") #The Bio_env and bv_step_ext routines below comes from http://menugget.blogspot.com.au/2011/06/clarke-and-ainsworths-bioenv-and-bvstep.html
```

Below is the output from an Rmarkdown R script for analyses of waterbird communities of the Warden and Gore wetland systems surveyed between Oct 2006 and Feb 2019

###Multi-variate analysis of Warden and Gore combined waterbird data by season. This data includes aerial counts for Warden system wetlands for 2006 to 2009 where ground counts not undertaken. These include Neridup and Bandy Creek (Oct 2006- Nov 2008), Ewans (Oct 2006) and Ewans and Mullet combined (Nov 2008 - have to ignore Nov 2008 Mullet ground counts), North Wheatfield (Oct 2006, Oct 2007, Nov 2008),  Gore data is included only for November 2009 onwards when ground counts are available. Pink Lake is excluded from analyses because it wasnt consistently surveyed. Birds counted only while in air and not allocated to any wetland are given the site code 'In flight'.
```{r}
  input.data <- read.csv("Warden_Gore_by_suite_06_19.csv") #csv file is a flat file of all species records by survey and system and survey method
  WG.bySeason <- input.data #to keep raw data as an intact file for later use
  WG.bySeason <- WG.bySeason[-grep("Unidentified", WG.bySeason$CommonName), ] #remove rows where birds not identified
  WG.bySeason <- WG.bySeason[-grep("Grassbird", WG.bySeason$CommonName), ] #remove records of little grassbird
  WG.bySeason <- WG.bySeason[!(WG.bySeason$Date == "23/02/2010" & WG.bySeason$SiteCode == "WRP004"), ] #remove duplicate data for Mullet Feb 2010
  WG.bySeason$Survey <- format(as.Date(WG.bySeason$Date), "%b-%Y") #create survey period column
  WG.bySeason <- aggregate(data=WG.bySeason, Count ~ CommonName + Survey, FUN="sum") #sum data by common name and survey period
  WG.bySeason$Count <- sqrt(WG.bySeason$Count) #convert counts to square root
```


```{r}
str(WG.bySeason)
WG.bySeason$Date <- format(as.Date(WG.bySeason$Date), "%b-%Y")
```



```{r}
  WG.bySeason <- melt(WG.bySeason) #melt data 
  WG.bySeason.m <- dcast(WG.bySeason, Survey ~ Common_Name, fill="0") #create matrix from flat-file
  rownames(WG.bySeason.m) <- WG.bySeason.m [,1] #use first column (season) as row.names
  WG.bySeason.m [,1] <- NULL #remove season column
  WG.bySeason.m <- data.matrix(WG.bySeason.m, rownames.force = NA) #convert data to numeric format (required for metaMDS)
```

```{r}
WG.MDS.bySeas <- metaMDS(WG.bySeason.m, distance = "bray")
```

MDS unconstrained ordination using 'metaMDS' in package Vegan to analyse effects of season irrespective of system.
```{r, include=TRUE}
WG.MDS.bySeas
```

```{r}
surv.seas<- unique(input.data[c("Survey", "Season")]) #extract a unique list of survey codes (mmm-yy) and season so can later allocate a season to each survey in the MDS output
  WG.MDS.bySeas.pts <- as.data.frame(WG.MDS.bySeas$points) #extract ordination coordinates (points)
  WG.MDS.bySeas.pts$Survey <- row.names(WG.MDS.bySeas.pts) #use survey period as row.names
  WG.MDS.bySeas.pts$Season <- surv.seas[match(WG.MDS.bySeas.pts$Survey, surv.seas$Survey), 2] #adds 'season' (spring versus summer) field to points by matching survey between points and surv.seas
```
<br>
  
Ordination plot of Warden and Gore data combined by season, stress = `r WG.MDS.bySeas$stress`  
```{r, include=TRUE}
  ggplot(WG.MDS.bySeas.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.4, 0.5) + ylim(-0.3, 0.3) + geom_point(aes(colour=Season), size=5) + coord_fixed(ratio = 1) + geom_text(aes(label=Survey), size=3, vjust=1, hjust=1.3)
```

```{r}
  WG.bySeas.Simper <- simper(WG.bySeason.m, surv.seas$Season) #Simper analysis to determine which species best correlated with differences between seasons of whole dataset
  summary(WG.bySeas.Simper)
```

```{r extract-top-8-species -from simper-for-contrasting-seasons}
  WG.Simper.ext <- WG.bySeas.Simper$Spring_Summer #extract results from Simper analysis
  WG.Sim.Average <- as.data.frame(WG.Simper.ext$average) #extract the average difference data
  WG.Sim.Spec <- as.data.frame(WG.Simper.ext$species) #extract species names
  WG.bySeas.Simper <- cbind(WG.Sim.Spec, WG.Sim.Average) #combine species with averages
  WG.bySeas.Simper <- WG.bySeas.Simper[ order(-WG.bySeas.Simper[, 2]), ] #order species and averages by average
colnames(WG.bySeas.Simper) = c("Species", "cont") #add new column names
  WG.bySeas.Simper <- as.vector(WG.bySeas.Simper[1:8, 1]) #extracts 
  WG.bySeas.Simper <- gsub("\\.", " ", WG.bySeas.Simper)
```

```{r}
  agg.spec.data <- aggregate(data=input.data, Count ~ Common_Name + Season + Survey, FUN="sum") #aggregate counts by common name, season and survey
  agg.spec.data <- as.data.frame(sapply(agg.spec.data, gsub, pattern = "-", replacement = " ")) #replace dashes with spaces
```

```{r extract-counts-of-top-8-species-from-simper-for-contrasting-system}
  agg.spec.data.8 <- agg.spec.data[agg.spec.data$Common_Name %in% WG.bySeas.Simper, ]
  rownames(agg.spec.data.8) <- NULL
  agg.spec.data.8 <- as.data.frame(agg.spec.data.8[, -3])
  agg.spec.data.8$Common_Name <- as.character(agg.spec.data.8$Common_Name)
```

```{r write-then-read-top-8-species-data-to-convert-to-numeric-for-analysis-by-season}
  write.csv(agg.spec.data.8, "agg.spec.data.8.csv")
  agg.spec.data.8 <- read.csv("agg.spec.data.8.csv")
```

Box plots of abundance in spring and summer for top eight species from simper analysis  
```{r, include=TRUE, facet-wrapped-box-plots-of-the-top-8-simper-species-by-season}
  qplot(x=Season, y=Count, data=agg.spec.data.8, geom="boxplot") + facet_wrap(~ Common_Name, scales="free_y", ncol=2)
```


####ORDINATION OF GORE SYSTEM VERSUS WARDEN SYSTEM

```{r import-reduce-aggregate-data}
  input.data2 <- read.csv("Warden and Gore ground counts by wetland suite 2006-2015.csv")
  WG.bySystem <- input.data2
  WG.bySystem <- WG.bySystem[WG.bySystem$WA_Fauna_Code_Name!="Wetland surveyed but no birds observed",]
  WG.bySystem <- WG.bySystem[-grep("Unidentified", WG.bySystem$Common_Name), ] #remove data for birds not identified to species
  WG.bySystem$Count <- sqrt(WG.bySystem$Count)
  WG.bySystem$Sys_Sur <- paste(WG.bySystem$System, WG.bySystem$Survey, sep = " ") # combine System and date to create one 'sample' variable for mds - i.e. combine system with season
  WG.bySystem <- aggregate(data=WG.bySystem, Count ~ Common_Name + Sys_Sur, FUN="sum")
```

```{r}
  WG.bySystem <- melt(WG.bySystem) #melt data
  WG.bySystem.m <- dcast(WG.bySystem, Sys_Sur ~ Common_Name, fill="0") #recast as matrix
rownames(WG.bySystem.m) <- WG.bySystem.m[,1]
  WG.bySystem.m[,1] <- NULL
  WG.bySystem.m <- data.matrix(WG.bySystem.m, rownames.force = NA) #convert data to numeric format (required for metaMDS)
```

MDS unconstrained ordination using metaMDS in package vegan to compare Warden and Gore wetland systems
```{r}
WG.MDS.bySyst <- metaMDS(WG.bySystem.m, distance = "bray")
```

```{r, include=TRUE}
WG.MDS.bySyst
```

```{r extract-coordinates-from-nMDS-and-add-season-to-resulting-data.frame-for Warden-versus-Gore-analysis}
system <- unique(input.data2$System) #extract a unique list of survey codes (mmm-yy) and season
WG.MDS.bySyst.pts <- as.data.frame(WG.MDS.bySyst$points) #extract coordinates from metaMDS
WG.MDS.bySyst.pts$Sys_Sur <- row.names(WG.MDS.bySyst.pts) #use row.names to create new System+Survey period variable
WG.MDS.bySyst.pts$Survey <- str_sub(WG.MDS.bySyst.pts$Sys_Sur, -6, -1) #extract Survey period from Sys-Sur variable
WG.MDS.bySyst.pts$System <- substr(WG.MDS.bySyst.pts$Sys_Sur, 1, 4) #extract System from Sur-sys variable
WG.MDS.bySyst.pts$System <- gsub("Ward", "Warden", WG.MDS.bySyst.pts$System)
```

ordination plot with symbols coloured by system (Warden versus Gore), `r WG.MDS.bySyst$stress`
```{r, include=TRUE}
ggplot(WG.MDS.bySyst.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.6, 0.5) + ylim(-0.3, 0.4) + geom_point(aes(colour=System), size=5) + geom_text(aes(label=Survey), size=3, vjust=1, hjust=1.3)
```

```{r}
  WG.bySyst.Sim <- simper(WG.bySystem.m, system) #Simper analysis to determine which species best correlated with differences between seasons of whole dataset
  summary(WG.bySyst.Sim)
```

```{r}
  WG.bySyst.Sim.ext <- WG.bySyst.Sim$Warden_Gore #extract name and contribution data from simper analysis
  WG.Sim.average <- as.data.frame(WG.bySyst.Sim.ext$average)#extract average contribution values
  WG.Sim.spec <- as.data.frame(WG.bySyst.Sim.ext$species)#extract species names
  WG.bySyst.Sim.ext <- cbind(WG.Sim.spec, WG.Sim.average) #combine average with species name
  WG.bySyst.Sim.ext <- WG.bySyst.Sim.ext[ order(-WG.bySyst.Sim.ext[,2]), ] #order by average value
colnames(WG.bySyst.Sim.ext) = c("Species", "cont") #add new column names
  WG.bySyst.Sim.ext <- as.vector(WG.bySyst.Sim.ext[1:8, 1]) #extract top eight species: first 8 rows
  WG.bySyst.Sim.ext <- gsub("\\.", " ", WG.bySyst.Sim.ext)
```

```{r}
agg.spec.data.2 <- aggregate(data=input.data2, Count ~ Common_Name + Survey + System, FUN="sum") #aggregare count data by system and survey
agg.spec.data.2 <- as.data.frame(sapply(agg.spec.data.2, gsub, pattern = "-", replacement = " ")) #replace dashes in species names with spaces to match simper output
agg.spec.data.2 <- agg.spec.data.2[agg.spec.data.2$Common_Name %in% WG.bySyst.Sim.ext, ] #restrict aggregated counts to just those top 8 species from simper
rownames(agg.spec.data.2) <- NULL
agg.spec.data.2$Common_Name <- as.character(agg.spec.data.2$Common_Name)
write.csv(agg.spec.data.2, "agg.spec.data.2.csv")
agg.spec.data.2 <- read.csv("agg.spec.data.2.csv")
```

Box plots of abundance in Warden and Gore for top eight species from simper analysis  
```{r, include=TRUE}
qplot(x=System, y=Count, data=agg.spec.data.2, geom="boxplot") + facet_wrap(~ Common_Name, scales="free_y", ncol=2) #produce facet wrapped box plots showing differences in top 8 simper species by wetland system
```

####WARDEN SYSTEM ANALYSIS ONLY<p/>

```{r}
#specify season to be worked on: "Spring" or "Summer"
Season <- "Summer"
```

```{r}
input.data3 <- read.csv("Warden and Gore ground counts by wetland suite 2006-2015.csv")
Warden <- input.data3
Warden <- Warden[Warden$WA_Fauna_Code_Name!="Wetland surveyed but no birds observed",] #remove zero counts
Warden <- Warden[-grep("Unidentified", Warden$Common_Name), ] #remove partial identifications
Warden$Count <- sqrt(Warden$Count) #square root data
Warden <- Warden[Warden$System=="Warden", ] #cut down to just Warden wetlands
Warden <- Warden[Warden$Season==Season, ] #cut down to just one season, as per Value Season above
Warden <- aggregate(data=Warden, Count ~ Common_Name + Survey, FUN="sum") #aggregate count data by species and survey
```

```{r convert-Warden-only-data-to-matrix}
Warden.m <- melt(Warden)
Warden.m <- dcast(Warden.m, Survey ~ Common_Name, fill="0")
rownames(Warden.m) <- Warden.m[,1]
Warden.m[,1] <- NULL
write.csv(Warden.m, "Warden_m.csv")
Warden.m <- read.csv("Warden_m.csv", row.names=1)
```

MDS unconstrained ordination using metaMDS in package vegan to analyse effects of season within the Warden system
```{r}
Warden.MDS <- metaMDS(Warden.m, distance = "bray")
```

```{r, include=TRUE}
Warden.MDS
```

```{r}
Warden.pts <- as.data.frame(Warden.MDS$points) #extract coordinates from ordination
Warden.pts$Survey <- row.names(Warden.pts)
#W.points$season <- c("spring","summer","summer","summer","summer","summer","summer","summer","spring","spring","spring","spring","spring","spring","spring","spring") # add season labels as new column in dataframe
```

```{r}
av.depths <- read.csv("average depths.csv") #load data file with average depths (per survey) for gauged Warden wetlands
av.depths <- av.depths[av.depths$Season == Season, ] #restrict to one season
Warden.pts$depth <- av.depths$depth #create new field for depth in the file with mds coordinates
Warden.pts$depth <- as.numeric(Warden.pts$depth)^3+30
```

Ordination graph of Warden system waterbird communities by surveys undertaken in spring, `r Warden.MDS$stress`  
```{r, include=TRUE}
#note, need to restrict data to just spring, see first code chunk in this section
ggplot(Warden.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.3, 0.5) + ylim(-0.35, 0.3) + coord_fixed(ratio = 1) + geom_point(shape=21, size=7, alpha=0.5, fill="blue", stroke=2) + geom_text(aes(label = Survey), vjust=2, hjust=1) + theme(legend.position = "none")
```

Ordination graph of Warden system waterbird communities by survey with surveys undertaken in spring scaled and coloured by average depth of gauged wetlands  
```{r, include=TRUE}
# note need to restrict both data matrix and average depths to just spring surveys
ggplot(Warden.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.3, 0.5) + ylim(-0.35, 0.3) + coord_fixed(ratio = 1) + geom_point(aes(size = depth, colour=depth)) + scale_size(range = c(3, 12)) + geom_text(aes(label = Survey), vjust=2, hjust=1) + theme(legend.position = "none")
```

Ordination graph of Warden system waterbird communities by surveys undertaken in summer  
```{r, include=TRUE}
# note, need to restrict data to just summer, see first code chunk in this section
ggplot(Warden.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.3, 0.3) + ylim(-0.25, 0.4) + coord_fixed(ratio = 1) + geom_point(shape=21, size=7, alpha=0.5, fill="blue", stroke=2) + geom_text(aes(label = Survey), vjust=2, hjust=1.5) + theme(legend.position = "none")
```

Ordination graph of Warden system waterbird communities by survey with surveys undertaken in summer scaled and coloured by average depth of gauged wetlands  
```{r, include=TRUE}
# note need to restrict both data matrix and average depths to just summer surveys
ggplot(Warden.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.3, 0.3) + ylim(-0.25, 0.3) + coord_fixed(ratio = 1) + geom_point(aes(size = depth, colour= depth)) + scale_size(range = c(3, 12)) + geom_text(aes(label = Survey), vjust=2, hjust=1.5) + theme(legend.position = "none")
```
