---
title: "Warden and Gore waterbird analyses"
author: "Adrian Pinder"
date and time: '`r Sys.time()`'
output: html_document
software: 'RStudio: Version 1.1.463 – © 2009-2018 RStudio, Inc. R version: `r getRversion()`'
editor_options: 
  chunk_output_type: console
---
  
Git repository https://github.com/AdrianMP62/Warden-2019  

RStudio: Version 1.1.463 – © 2009-2018 RStudio, Inc. R version: `r getRversion()`  
Date and time: `r Sys.time()`

Uses the following datafiles:  
*  Warden_Gore_by_suite_06_19.csv (System,	SurveyType,	SiteName,	SiteCode,	Subsite	Date,	CommonName,	SpeciesCode,	Count) (e.g. Warden,	aerial,	Neridup Suite,	WRP001,	combined,	11/10/2006,	Great Egret,	greg,	1)
*  Average depths.csv (average depths of Warden wetlands for each survey period).
*  Taxonomy (CommonName, Order and informal group name in consecutive columns. CommonNames match those in rden_Gore_by_suite_06_19.csv)

Date and time: `r Sys.time()`

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(include = FALSE)
knitr::opts_knit$set(root.dir = '../') 
```

```{r}
options(scipen=999)
```

```{r set-seed}
set.seed(1234)
```

```{r load-packages, results="hide", warning=FALSE, message=FALSE}
Packages <- c("vegan", "ggrepel", "ggplot2", "lubridate", "stringr", "gsubfn", "reshape2", "knitr", "simba", "zoo", "car")
lapply(Packages, library, character.only = TRUE)
#source("./functions/bio_env_ext.R") #The Bio_env and bv_step_ext routines below comes from http://menugget.blogspot.com.au/2011/06/clarke-and-ainsworths-bioenv-and-bvstep.html
```

#Analyses of waterbird data from the Warden and Gore Ramsar wetlands 2006 to 2019
  
The following is the output from an Rmarkdown R script for analyses of waterbird communities of the Warden and Gore wetland systems surveyed between Oct 2006 and Feb 2019.  
  
These analyses are based primarily on ground counts. Aerial counts are only for a few Warden system wetlands in 2006 to 2009 where ground counts were not undertaken. These include Neridup and Bandy Creek (Oct 2006-Nov 2008), Ewans (Oct 2006) and Ewans and Mullet combined (Nov 2008 - have to ignore Nov 2008 Mullet ground counts), North Wheatfield (Oct 2006, Oct 2007, Nov 2008). Note that the Nov 2008 aerial data was collected two weeks prior to the ground counts.  Gore data is included only for November 2009 onwards when ground counts are available. Pink Lake is excluded from analyses because it wasnt consistently surveyed. Birds counted only while in air and not allocated to any particular wetland are given the site code 'In flight'. These conditions apply to all following analyses.  
  
##Multi-variate analysis of Warden and Gore combined waterbird data by season  

This ordination shows that, by and large, with data from systems (Warden and Gore) combined, there is a difference between waterbird communities present in spring and those present in late summer.

```{r import-and-restructure-Warden-and-Gore-waterbird-data}
input.data <- read.csv("./raw_data/Warden_Gore_by_suite_06_19.csv")
 WG.bySeason <- input.data
  WG.bySeason <- WG.bySeason[-grep("Unidentified", WG.bySeason$CommonName), ] #remove rows where birds not identified
  WG.bySeason <- WG.bySeason[-grep("Grassbird", WG.bySeason$CommonName), ] #remove records of little grassbird
  WG.bySeason <- WG.bySeason[!(WG.bySeason$Date == "23/02/2010" & WG.bySeason$SiteCode == "WRP004"), ] #remove duplicate data for Mullet Feb 2010
  WG.bySeason$Date <- as.Date(WG.bySeason$Date, format="%d/%m/%Y")
  WG.bySeason$Survey <- WG.bySeason$Date #create Survey field from Date
  WG.bySeason$Survey <- format(as.Date(WG.bySeason$Survey), "%b-%Y") #change format to mmm-YYYY
  WG.bySeason$Survey <- gsub("Oct-2018", "Nov-2018", WG.bySeason$Survey) #lump survey dates that straddle Oct/Nov 2018
  yq <- as.yearqtr(as.yearmon(WG.bySeason$Date, "%Y/%m/%d"))
WG.bySeason$Season <- factor(format(yq, "%q"), levels = 1:4, labels = c("summer", "autumn", "winter", "spring")) #create Season (spring vs summer) variable
write.csv(WG.bySeason, "./analysis_data/WG.bySeason.csv")
```

```{r aggregare-by-survey-and-species}
  WG.bySeason.agg <- aggregate(data=WG.bySeason, Count ~ CommonName + Survey + Season, FUN="sum") #sum data by common name and survey period
  WG.bySeason.agg$Count <- sqrt(WG.bySeason.agg$Count) #convert counts to square root
```

```{r create-matrix-for-ordination}
  WG.bySeason.agg <- melt(WG.bySeason.agg) #melt data 
  WG.bySeason.m <- dcast(WG.bySeason.agg, Survey ~ CommonName, fill="0") #create matrix from flat-file
  rownames(WG.bySeason.m) <- WG.bySeason.m [,1] #use first column (season) as row.names
  WG.bySeason.m [,1] <- NULL #remove season column
  WG.bySeason.m <- data.matrix(WG.bySeason.m, rownames.force = NA) #convert data to numeric format (required for metaMDS)
```

```{r undertake-2D-and-3D_ordination}
WG.MDS.bySeas <- metaMDS(WG.bySeason.m, distance = "bray")
WG.3dMDS.bySeas <- metaMDS(WG.bySeason.m, distance = "bray", k=3)
```

```{r extract-list-of-seasons-for-ordination-plots}
surv.seas<- unique(WG.bySeason[c("Survey", "Season")]) #extract a unique list of survey codes (mmm-yy) and season so can later allocate a season to each survey in the MDS output
surv.seas <- surv.seas[order(surv.seas$Survey), ] #reorder alphabetically by Survey (Dec-2011 to Oct-2012)
```
  
```{r create-2D-dataset-for-plotting}
  WG.MDS.bySeas.pts <- as.data.frame(WG.MDS.bySeas$points) #extract 2D ordination coordinates (points)
  WG.MDS.bySeas.pts$Survey <- row.names(WG.MDS.bySeas.pts) #use survey period as row.names
  WG.MDS.bySeas.pts$Season <- surv.seas[match(WG.MDS.bySeas.pts$Survey, surv.seas$Survey), 2] #adds 'season' (spring versus summer) (from column '2' of Surv.Seas) to ordination points by matching survey between points and surv.seas
```

```{r create-3D-dataset-for-plotting}
  WG.3dMDS.bySeas.pts <- as.data.frame(WG.3dMDS.bySeas$points) #extract 3D ordination coordinates (points)
  WG.3dMDS.bySeas.pts$Survey <- row.names(WG.3dMDS.bySeas.pts) #use survey period as row.names
  WG.3dMDS.bySeas.pts$Season <- surv.seas[match(WG.3dMDS.bySeas.pts$Survey, surv.seas$Survey), 2] #adds 'season' (spring versus summer) (from column '2' of Surv.Seas) to ordination points by matching survey between points and surv.seas
```
 
2D ordination plot of Warden and Gore data combined by season, stress = `r WG.MDS.bySeas$stress`.
```{r, 2D-ordination-plot-by-season, include=TRUE}
ggplot(WG.MDS.bySeas.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.4, 0.35) + ylim(-0.3, 0.2) + geom_point(aes(colour=Season), size=5) + coord_fixed(ratio = 1) + geom_text(aes(label=Survey), size=3, vjust=1, hjust=1.3)
```

3D ordination plot of Warden and Gore data combined by season, stress = `r WG.MDS.bySeas$stress`.
```{r, 3D-ordination-plot-by-season, include=TRUE}
ggplot(WG.3dMDS.bySeas.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.4, 0.35) + ylim(-0.3, 0.3) + geom_point(aes(colour=Season, shape=System), size=5) + coord_fixed(ratio = 1) + geom_text(aes(label=Survey), size=3, vjust=1, hjust=1.3) + theme (plot.margin=unit(c(0,0,0,5),"mm"))
```

```{r run-simper-analysis-to-determine-influence-of-species-by-season}
  WG.bySeas.Simper <- simper(WG.bySeason.m, surv.seas$Season) #Simper analysis to determine which species best correlated with differences between seasons of whole dataset
```

```{r extract-top-8-spp-from-simper-to-contrast-seasons}
  WG.Simper.ext <- WG.bySeas.Simper$spring_summer #extract results from Simper analysis
  WG.Sim.Average <- as.data.frame(WG.Simper.ext$average) #extract the average difference data
  WG.Sim.Spec <- as.data.frame(WG.Simper.ext$species) #extract species names
  WG.bySeas.Simper2 <- cbind(WG.Sim.Spec, WG.Sim.Average) #combine species with averages
  WG.bySeas.Simper2 <- WG.bySeas.Simper2[ order(-WG.bySeas.Simper2[, 2]), ] #order species and averages by average
colnames(WG.bySeas.Simper2) = c("Species", "cont") #add new column names
  WG.bySeas.Simper2 <- as.vector(WG.bySeas.Simper2[1:8, 1]) #extracts 
```

```{r aggregate-by-sppname-and-survey-for season-simper}
  agg.spec.data <- aggregate(data=WG.bySeason, Count ~ CommonName + Season + Survey, FUN="sum") #aggregate counts by common name, season and survey
  #agg.spec.data <- as.data.frame(sapply(agg.spec.data, gsub, pattern = "-", replacement = " ")) #replace dashes with spaces
```

```{r extract-counts-of-top-8-species-from-simper-for-contrasting-system}
  agg.spec.data.8 <- agg.spec.data[agg.spec.data$CommonName %in% WG.bySeas.Simper2, ]
  rownames(agg.spec.data.8) <- NULL #reset row names
  agg.spec.data.8 <- as.data.frame(agg.spec.data.8[, -3]) #remove Survey field
  agg.spec.data.8$CommonName <- as.character(agg.spec.data.8$CommonName)
```

```{r write-then-read-top-8-species-data-to-convert-to-numeric-for-analysis-by-season}
  write.csv(agg.spec.data.8, "agg.spec.data.8.csv")
  agg.spec.data.8 <- read.csv("agg.spec.data.8.csv")
```

These box plots show the abundance of the top 8 species contributiong to waterbird communities being different in spring and summer (Gore and Warden systems combined).  
```{r simper-by-season-box-plots, include=TRUE}
  qplot(x=Season, y=Count, data=agg.spec.data.8, geom="boxplot") + facet_wrap(~ CommonName, scales="free_y", ncol=2)
```


##ORDINATION OF GORE SYSTEM VERSUS WARDEN SYSTEM  

The ordination plot below shows that the Warden and Gore wetland systems consistently support different waterbird communities, i.e. for waterbirds they have different conservation values. It can also be seen that there is greater seasonal differences in the Warden system than in the Gore system (i.e. greater overlap between positions of the summer and spring surveys amongst the Gore surveys).  

the Nov 2018 and Feb 2019 surveys are within the range of community compositions surveyed since 2006 for both systems, despite the very high counts in Feb 2019 and low Warden count in Nov 2018.

```{r import-reduce-aggregate-data}
 input.data2 <- read.csv("./raw_data/Warden_Gore_by_suite_06_19.csv")
  WG.bySystem <- input.data2
    WG.bySystem <- WG.bySystem[-grep("Unidentified", WG.bySystem$CommonName), ] #remove rows where birds not identified
  WG.bySystem <- WG.bySystem[-grep("Grassbird", WG.bySystem$CommonName), ] #remove records of little grassbird
  WG.bySystem <- WG.bySystem[!(WG.bySystem$Date == "23/02/2010" & WG.bySystem$SiteCode == "WRP004"), ] #remove duplicate data for Mullet Feb 2010
  WG.bySystem$Date <- as.Date(WG.bySystem$Date, format="%d/%m/%Y")
  WG.bySystem$Survey <- WG.bySystem$Date #create Survey field from Date
  WG.bySystem$Survey <- format(as.Date(WG.bySystem$Survey), "%b-%Y") #change format to mmm-YYYY
  WG.bySystem$Survey <- gsub("Oct-2018", "Nov-2018", WG.bySystem$Survey) #lump survey dates that straddle Oct/Nov 2018
  yq <- as.yearqtr(as.yearmon(WG.bySystem$Date, "%Y/%m/%d"))
WG.bySystem$Season <- factor(format(yq, "%q"), levels = 1:4, labels = c("summer", "autumn", "winter", "spring")) #create Season (spring vs summer) variable
  WG.bySystem$Count <- sqrt(WG.bySystem$Count)
  WG.bySystem$Sys_Sur <- paste(WG.bySystem$System, WG.bySystem$Survey, sep = " ") # combine System and date to create one 'sample' variable for mds - i.e. combine system with season
write.csv(WG.bySystem, "./WG.bySystem.csv")
```

```{r aggregate-by-Common-Name}
WG.bySystem.agg <- aggregate(data=WG.bySystem, Count ~ CommonName + Sys_Sur, FUN="sum")
WG.bySystem.agg$Count <- sqrt(as.numeric(WG.bySystem.agg$Count))
```

```{r create_matrix_for_ordaniation}
  WG.bySystem.agg <- melt(WG.bySystem.agg) #melt data
  WG.bySystem.m <- dcast(WG.bySystem.agg, Sys_Sur ~ CommonName, fill="0") #recast as matrix
rownames(WG.bySystem.m) <- WG.bySystem.m[,1]
  WG.bySystem.m[,1] <- NULL
  WG.bySystem.m <- data.matrix(WG.bySystem.m, rownames.force = NA) #convert data to numeric format (required for metaMDS)
```

MDS unconstrained ordination using metaMDS in package vegan to compare Warden and Gore wetland systems
```{r ordination-by-system}
WG.MDS.bySyst <- metaMDS(WG.bySystem.m, distance = "bray")
```

```{r extract-coordinates-from-nMDS-and-add-season-to-resulting-data.frame-for-Warden-versus-Gore-analysis}
system <- unique(input.data2$System) #extract a unique list of systems (Gore, Warden)
WG.MDS.bySyst.pts <- as.data.frame(WG.MDS.bySyst$points) #extract coordinates from metaMDS
WG.MDS.bySyst.pts$Sys_Sur <- row.names(WG.MDS.bySyst.pts) #use row.names to create new System+Survey period variable
WG.MDS.bySyst.pts$Survey <- str_sub(WG.MDS.bySyst.pts$Sys_Sur, -8, -1) #extract Survey period from Sys-Sur variable by extracting last 8 characters
WG.MDS.bySyst.pts$System <- substr(WG.MDS.bySyst.pts$Sys_Sur, 1, 4) #extract System from Sur-sys variable by extracting first 4 variables
WG.MDS.bySyst.pts$System <- gsub("Ward", "Warden", WG.MDS.bySyst.pts$System) #convert 'Ward' to 'Warden'
#WG.MDS.bySyst.pts <- WG.MDS.bySyst.pts[with(WG.MDS.bySyst.pts, order(Survey)), ] #order by survey
WG.MDS.bySyst.pts$Season <-  substr(WG.MDS.bySyst.pts$Survey, 1, 3) #extract months
WG.MDS.bySyst.pts$Season <- recode(WG.MDS.bySyst.pts$Season, "'Feb'='summer';c('Dec','Oct','Nov')='spring'") #convert months to season
WG.MDS.bySyst.pts$SystSeas <- paste(WG.MDS.bySyst.pts$System, WG.MDS.bySyst.pts$Season, sep=" in ")
str(WG.MDS.bySyst.pts) #join System and season
```

ordination plot with symbols coloured by system (Warden versus Gore), `r WG.MDS.bySyst$stress`
```{r plot-2D-ordination-by-system, include=TRUE}
ggplot(WG.MDS.bySyst.pts, aes(x=MDS1, y=MDS2, shape=SystSeas, colour=SystSeas)) + xlim(-0.45, 0.3) + ylim(-0.3, 0.25) + coord_fixed(ratio = 1) + geom_point(size=5) + scale_color_manual(values=c("blue","blue","red","red")) + scale_shape_manual(values=c(0,15,2,17)) + geom_text(aes(label=Survey), colour="black", size=3, vjust=1, hjust=1.3)
```

However, the stress level of the above two-dimensional ordintion is unacceptably high at `r WG.MDS.bySyst$stress`. A three-dimensional ordination was therefore undertaken.

3D MDS unconstrained ordination using metaMDS in package vegan to compare Warden and Gore wetland systems
```{r 3Dordination-by-system}
WG.3dMDS.bySyst <- metaMDS(WG.bySystem.m, distance = "bray", k=3)
plot(WG.3dMDS.bySyst, xlim=c(-1,1), display = c("sites", "species"), type="t")
```

```{r extract-coordinates-from-3D-nMDS-and-add-season-to-resulting-data.frame-for Warden-versus-Gore-analysis}
system <- unique(input.data2$System) #extract a unique list of systems (Gore, Warden)
WG.3dMDS.bySyst.pts <- as.data.frame(WG.3dMDS.bySyst$points) #extract coordinates from metaMDS
WG.3dMDS.bySyst.pts$Sys_Sur <- row.names(WG.3dMDS.bySyst.pts) #use row.names to create new System+Survey period variable
WG.3dMDS.bySyst.pts$Survey <- str_sub(WG.3dMDS.bySyst.pts$Sys_Sur, -8, -1) #extract Survey period from Sys-Sur variable by extracting last 8 characters
WG.3dMDS.bySyst.pts$System <- substr(WG.3dMDS.bySyst.pts$Sys_Sur, 1, 4) #extract System from Sur-sys variable by extracting first 4 variables
WG.3dMDS.bySyst.pts$System <- gsub("Ward", "Warden", WG.3dMDS.bySyst.pts$System) #convert 'Ward' to 'Warden'
#WG.3dMDS.bySyst.pts <- WG.3dMDS.bySyst.pts[with(WG.3dMDS.bySyst.pts, order(Survey)), ] #order by survey
WG.3dMDS.bySyst.pts$Season <-  substr(WG.3dMDS.bySyst.pts$Survey, 1, 3) #extract months
WG.3dMDS.bySyst.pts$Season <- recode(WG.3dMDS.bySyst.pts$Season, "'Feb'='summer';c('Dec','Oct','Nov')='spring'") #convert months to season
WG.3dMDS.bySyst.pts$SystSeas <- paste(WG.3dMDS.bySyst.pts$System, WG.3dMDS.bySyst.pts$Season, sep=" in ")
str(WG.3dMDS.bySyst.pts) #join System and season
```

```{r plot-3D-ordination-by-system, include=TRUE}
ggplot(WG.3dMDS.bySyst.pts, aes(x=MDS1, y=MDS2, shape=SystSeas, colour=SystSeas)) + xlim(-0.5, 0.4) + ylim(-0.3, 0.3) + coord_fixed(ratio = 1) + geom_point(size=5) + scale_color_manual(values=c("blue","blue","red","red")) + scale_shape_manual(values=c(0,15,2,17)) + theme(legend.title = element_blank()) + theme(legend.position = "top") + theme(legend.text=element_text(size=12)) + theme(plot.margin=unit(c(0,0,0,0),"mm")) + geom_text_repel(aes(label=Survey), colour="grey", size=5, vjust=1, hjust=1.5)
```

```{r simper-analysis-of-species-separating-the-two-systems}
  WG.bySyst.Sim <- simper(WG.bySystem.m, WG.3dMDS.bySyst.pts$System) #Simper analysis to determine which species best correlated with differences between seasons of whole dataset
summary(WG.bySyst.Sim, ordered=TRUE)
```

```{r extract-simperdata-for-top10pp}
  WG.bySyst.Sim.ext <- WG.bySyst.Sim$Gore_Warden #extract name and contribution data from simper analysis
  WG.bySyst.Sim.average <- as.data.frame(WG.bySyst.Sim.ext$average)#extract average contribution values
  WG.bySyst.Sim.spec <- as.data.frame(WG.bySyst.Sim.ext$species)#extract species names
  WG.bySyst.Sim.ext <- cbind(WG.bySyst.Sim.spec, WG.bySyst.Sim.average) #combine average with species name
  WG.bySyst.Sim.ext <- WG.bySyst.Sim.ext[ order(-WG.bySyst.Sim.ext[,2]), ] #order by average value
colnames(WG.bySyst.Sim.ext) = c("Species", "cont") #add new column names
  WG.bySyst.Sim.ext <- as.vector(WG.bySyst.Sim.ext[1:10, 1]) #extract top eight species: first 8 rows
  #WG.bySyst.Sim.ext <- gsub("\\.", " ", WG.bySyst.Sim.ext)
```

```{r restrict-countdata-to-simper-top10spp}
WG.bySystem$Count2 <- WG.bySystem$Count^2
agg.spec.data.2 <- aggregate(data=WG.bySystem, Count2 ~ CommonName + Survey + System, FUN="sum") #aggregare count data by system and survey
#agg.spec.data.2 <- as.data.frame(sapply(agg.spec.data.2, gsub, pattern = "-", replacement = " ")) #replace dashes in species names with spaces to match simper output
agg.spec.data.2 <- agg.spec.data.2[agg.spec.data.2$CommonName %in% WG.bySyst.Sim.ext, ] #restrict aggregated counts to just those top 8 species from simper
rownames(agg.spec.data.2) <- NULL
agg.spec.data.2$CommonName <- as.character(agg.spec.data.2$CommonName)
colnames(agg.spec.data.2)[4] <- "Abundance"
write.csv(agg.spec.data.2, "agg.spec.data.2.csv")
agg.spec.data.2 <- read.csv("agg.spec.data.2.csv")
```

A simper analysis was undertaken to determine which species were most responsible for the separation of Warden and Gore wetland systems. For the top ten species, box plots of abundance on the Warden versus Gore systems are provided below.  

```{r boxplots-simper-top10spp, include=TRUE}
qplot(x=System, y=Abundance, data=agg.spec.data.2, geom="boxplot") + facet_wrap(~ CommonName, scales="free_y", ncol=2) + theme(axis.title.y = element_text(hjust=0.5)) #produce facet wrapped box plots showing differences in top 10 simper species by wetland system
```

undertake Mann-Whitney for each of the 10 top Simper species and combine results into file called ttest.results<p/>
```{r}
WG.bySyst.Sim.ext <- WG.bySyst.Sim.ext[-8]
WG.bySyst.Sim.ext
ttest.v <- vector()
for (i in 1:9) {
tt.data <- subset(agg.spec.data.2, CommonName == WG.bySyst.Sim.ext[i])
print(ttest <- wilcox.test(log10(Abundance) ~ System, data = tt.data))
ttest.v <- c(ttest.v, ttest$p.value)
}
ttest.v
ttest.results <- data.frame(CommonName=WG.bySyst.Sim.ext, p.value=ttest.v)
ttest.results
```


##WARDEN SYSTEM ANALYSIS ONLY  

###Ordinations and associated analyses
```{r input_data_and_prepare_for_just_Warden_ordinations}
input.data3 <- read.csv("./raw_data/Warden_Gore_by_suite_06_19.csv")
Warden <- input.data3
Warden <- Warden[-grep("Unidentified", Warden$CommonName), ] #remove rows where birds not identified
Warden <- Warden[Warden$System=="Warden", ] #cut down to just Warden wetlands
Warden <- Warden[-grep("Grassbird", Warden$CommonName), ] #remove records of little grassbird
  Warden <- Warden[!(Warden$Date == "23/02/2010" & Warden$SiteCode == "WRP004"), ] #remove duplicate data for Mullet Feb 2010
  Warden$Date <- as.Date(Warden$Date, format="%d/%m/%Y")
  Warden$Survey <- Warden$Date #create Survey field from Date
  Warden$Survey <- format(as.Date(Warden$Survey), "%b-%Y") #change format to mmm-YYYY
  Warden$Survey <- gsub("Oct-2018", "Nov-2018", Warden$Survey) #lump survey dates that straddle Oct/Nov 2018 and call them Nov-2018
  yq <- as.yearqtr(as.yearmon(Warden$Date, "%Y/%m/%d")) #create a temporary season field
Warden$Season <- factor(format(yq, "%q"), levels = 1:4, labels = c("summer", "autumn", "winter", "spring")) #create Season (spring vs summer) variable
  Warden.sum <- Warden[Warden$Season == "summer", ] #summer only data
  Warden.spr <- Warden[Warden$Season == "spring", ] #spring only data
```

```{r aggregate_count_data_by_Common_name_for_each_season}
Warden.sum.agg <- aggregate(data=Warden.sum, Count ~ CommonName + Survey, FUN="sum") #aggregate count data by species and survey
Warden.sum.agg$Count <- sqrt(Warden.sum.agg$Count)
Warden.spr.agg<- aggregate(data=Warden.spr, Count ~ CommonName + Survey, FUN="sum") #aggregate count data by species and survey
Warden.spr.agg$Count <- sqrt(Warden.spr.agg$Count)
```

```{r convert-Warden-only-data-to-seasonal-matrices}
#summer
Warden.sum.m <- melt(Warden.sum.agg)
Warden.sum.m <- dcast(Warden.sum.m, Survey ~ CommonName, fill="0")
rownames(Warden.sum.m) <- Warden.sum.m[,1]
Warden.sum.m[,1] <- NULL
write.csv(Warden.sum.m, "Warden_sum_m.csv")
Warden.sum.m <- read.csv("Warden_sum_m.csv", row.names=1)
#spring
Warden.spr.m <- melt(Warden.spr.agg)
Warden.spr.m <- dcast(Warden.spr.m, Survey ~ CommonName, fill="0")
rownames(Warden.spr.m) <- Warden.spr.m[,1]
Warden.spr.m[,1] <- NULL
write.csv(Warden.spr.m, "./analysis_data/Warden_spr_m.csv")
Warden.spr.m <- read.csv("./analysis_data/Warden_spr_m.csv", row.names=1)
```

MDS unconstrained ordination using metaMDS in package vegan to analyse effects of season within the Warden system
```{r 2D-ordinations-Warden-only}
Warden.sum.MDS <- metaMDS(Warden.sum.m, distance = "bray")
Warden.spr.MDS <- metaMDS(Warden.spr.m, distance = "bray")
```

```{r extract-Warden-coords-for-plot}
Warden.sum.MDS.pts <- as.data.frame(Warden.sum.MDS$points) #extract coordinates from ordination
Warden.sum.MDS.pts$Survey <- row.names(Warden.sum.MDS.pts)
Warden.spr.MDS.pts <- as.data.frame(Warden.spr.MDS$points) #extract coordinates from ordination
Warden.spr.MDS.pts$Survey <- row.names(Warden.spr.MDS.pts)
#W.points$season <- c("spring","summer","summer","summer","summer","summer","summer","summer","spring","spring","spring","spring","spring","spring","spring","spring") # add season labels as new column in dataframe
```

```{r-load-env-data-split-by-season}
av.depths <- read.csv("./raw_data/average depths.csv") #load data file with average depths (per survey) for gauged Warden wetlands
#for average depths only
av.depths.spr <- av.depths[av.depths$Season == "Spring", ] #restrict to spring
D.spr <- as.numeric(av.depths.spr$depth) #create numeric variable for spring depth
av.depths.sum <- av.depths[av.depths$Season == "Summer", ] #restrict to summer
D.sum <- as.numeric(av.depths.sum$depth) #create numeric variable for summer depth
#for multiple variables
BEdat.spr <- as.data.frame(av.depths.spr[, 3:5], row.names = as.character(av.depths.spr$Survey))
BEdat.sum <- as.data.frame(av.depths.sum[, 3:5], row.names = as.character(av.depths.sum$Survey))
```

Ordination graph of Warden system waterbird communities by survey with surveys undertaken in summer scaled and coloured by average depth of gauged wetlands  
```{r summer-Warden-ord-plot-bydepths, include=TRUE}
ggplot(Warden.sum.MDS.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.3, 0.3) + ylim(-0.2, 0.2) + coord_fixed(ratio = 1) + geom_point(aes(size = D.sum, colour=D.sum)) + scale_size(range = c(3, 12)) + geom_text(aes(label = Survey), cex=3, vjust=2, hjust=1.25) + theme(legend.position = "none")
```

Ordination graph of Warden system waterbird communities by survey with surveys undertaken in spring scaled and coloured by average depth of gauged wetlands  
```{r spring-Warden-ord-plot-bydepths, include=TRUE}
# note need to restrict both data matrix and average depths to just summer surveys
ggplot(Warden.spr.MDS.pts, aes(x=MDS1, y=MDS2)) + xlim(-0.3, 0.3) + ylim(-0.2, 0.25) + coord_fixed(ratio = 1) + geom_point(aes(size = D.spr, colour=D.spr)) + scale_size(range = c(3, 12)) + geom_text(aes(label = Survey), cex=3, vjust=2, hjust=1.25) + theme(legend.position = "none") + scale_x_reverse(limits=c(0.3, -0.3))
```

Bio-env analysis to determine influence of depth on waterbrd communities on the Warden wetlands
```{r bio-env-Warden}
#uses the multiple env variable dataset BEdat.spr and BEdat.sum rather than just the depth only datast\et.
D.spr.df <- as.data.frame(D.spr, row.names = row.names(Warden.spr.m)) #extract depths
Warden.spr.be <- bioenv(Warden.spr.m, BEdat.spr, method = "spearman", index = "bray")
Warden.spr.be
D.sum.df <- as.data.frame(D.sum, row.names = row.names(Warden.sum.m)) #extract depths
Warden.sum.be <- bioenv(Warden.sum.m, BEdat.sum, method = "spearman", index = "bray")
Warden.sum.be
```

###Stacked column plots for Warden system waterbirds

```{r input_data_and_prepare_for_just_Warden_stacked_column_graphs}
input.data3 <- read.csv("./raw_data/Warden_Gore_by_suite_06_19.csv")
Warden.stack <- input.data3
Warden.stack <- Warden.stack[Warden.stack$System=="Warden", ] #cut down to just Warden.stack wetlands
Warden.stack <- Warden.stack[-grep("Grassbird", Warden.stack$CommonName), ] #remove records of little grassbird
  Warden.stack <- Warden.stack[!(Warden.stack$Date == "23/02/2010" & Warden.stack$SiteCode == "WRP004"), ] #remove duplicate data for Mullet Feb 2010
  Warden.stack$Date <- as.Date(Warden.stack$Date, format="%d/%m/%Y")
  Warden.stack$Survey <- Warden.stack$Date #create Survey field from Date
  Warden.stack$Survey <- format(as.Date(Warden.stack$Survey), "%b-%Y") #change format to mmm-YYYY
  Warden.stack$Survey <- gsub("Oct-2018", "Nov-2018", Warden.stack$Survey) #lump survey dates that straddle Oct/Nov 2018 and call them Nov-2018
  yq <- as.yearqtr(as.yearmon(Warden.stack$Date, "%Y/%m/%d")) #create a temporary season field
Warden.stack$Season <- factor(format(yq, "%q"), levels = 1:4, labels = c("summer", "autumn", "winter", "spring"))
Warden.stack$SurvOrd <- as.character(Warden.stack$Survey)
Warden.stack$SurvOrd <- as.numeric(str_replace_all(Warden.stack$SurvOrd, c("Feb-2008" = "1", "Feb-2010" = "2", "Feb-2011" = "3","Feb-2012" = "4", "Feb-2013" = "5", "Feb-2014" = "6", "Feb-2015" = "7", "Feb-2019" = "8", "Oct-2006" = "1", "Oct-2007" = "2", "Nov-2008" = "3", "Nov-2009" = "4", "Nov-2010" = "5", "Dec-2011" = "6", "Oct-2012" = "7", "Nov-2013"= "8", "Nov-2014" = "9", "Nov-2018" = "10")))
```

```{r import-taxonomy-data-and-add-Group-to-Warden-data}
Taxonomy <- read.xlsx("./raw_data/Taxonomy.xlsx", sheetIndex=1)
Warden.stack$Group <- Taxonomy[match(Warden.stack$CommonName, Taxonomy$CommonName), 3]
```

```{r}
Warden.stack.agg <- aggregate(data=Warden.stack, Count ~ Group + SurvOrd + Survey + Season, FUN="sum") #aggregate count data by species and survey
```

```{r plot-spring-survey-abundances}
Warden.stack.spr <- Warden.stack.agg[Warden.stack.agg$Season=="spring", ] #cut down to just spring data
ggplot(Warden.stack.spr, aes(x=reorder(Survey, SurvOrd), y=Count, fill=Group)) + geom_bar(stat="identity", colour = "black") + scale_y_continuous(breaks = seq(0,30000, by=5000), limits=c(0,30000)) + labs(x="Survey", y="Abundance") + theme (axis.text.x=element_text(angle=90)) + theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))) + theme(axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0))) + scale_fill_manual(values = c("#CD9B9B","#912CEE","#104E8B","#698B22","#00EEEE","#A2B5CD","#FFA500","#FFE1FF","#B4CDCD","#C1FFc1","#A52A2A")) #uses Date to reorder survey on x axis
```

```{r plot-summer-survey-abundances}
Warden.stack.sum <- Warden.stack.agg[Warden.stack.agg$Season=="summer", ] #cut down to just summer data
ggplot(Warden.stack.sum, aes(x=reorder(Survey, SurvOrd), y=Count, fill=Group)) + geom_bar(stat="identity", colour = "black") + scale_y_continuous(breaks = seq(0,30000, by=5000), limits=c(0,30000)) + labs(x="Survey", y="Abundance") + theme (axis.text.x=element_text(angle=90)) + theme(axis.title.y = element_text(margin = margin(t = 0, r = 20, b = 0, l = 0))) + theme(axis.title.x = element_text(margin = margin(t = 20, r = 0, b = 0, l = 0))) + scale_fill_manual(values = c("#CD9B9B","#912CEE","#104E8B","#698B22","#00EEEE","#A2B5CD","#FFA500","#FFE1FF","#B4CDCD","#C1FFc1","#A52A2A")) #uses Date to reorder survey on x axis
```

### Esperance rainfall for Warden wetlands

```{r load-and-prepare-rainfall-data}
rainfall <- read.csv("./raw_data/multi station CDM.csv")
rainfall$Month <- as.yearmon(rainfall$Month, "%b-%y")
rainfall$Month.2 <- rainfall$Month
rainfall$Month.2 <- as.Date(rainfall$Month)
databreaks <- seq(as.Date("2004-01-01"), as.Date("2019-03-01"), by="3 month")
labels <- as.yearmon(databreaks, "%b-%y")
labels
```

```{r plot-of-rainfall-data}
ggplot(rainfall, aes(x = Month.2, y = CDMM, colour=Station)) + geom_line() + theme (axis.text.x=element_text(angle=90)) + scale_x_date(breaks=databreaks, labels = labels)
```

```{r calculate spring total waterbird-abundances}
Warden.spr$year <- lubridate::year(Warden.spr$Date)
Warden.spr.yearsums <- aggregate(data=Warden.spr, Count ~ year + year, FUN="sum")
write.csv(Warden.spr.yearsums, "./analysis_data/warden_spr_yearsums.csv")
```

```{r calculate summer total waterbird-abundances}
Warden.sum$year <- lubridate::year(Warden.sum$Date)
Warden.sum.yearsums <- aggregate(data=Warden.sum, Count ~ year + year, FUN="sum")
write.csv(Warden.sum.yearsums, "./analysis_data/warden_sum_yearsums.csv")
```